The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-07-15 VegOzone.tex
Spell checked. Added more to abstracted. Elaborated on introduction.

\begin{abstract}
  The summer 2018 and summer 2019 did not differ significantly in the level of ozone. On the other hand, the temperatures (precipitation and cloud cover/PAR?) differed quite a lot. We also observed ozone-induced visible injuries on clovers in the ozone garden at Svanhovd in 2018, but not in 2019. This may be due to differences in uptake of ozone ($\mathrm{POD_y}$) between the two years. Since DO3SE can be run on birch/beech and conifers, we would like to investigate whether the critical levels for these tree species were crossed during the two summer seasons. Ozone concentrations for the two missing weeks in July 2018 have been estimated based on the measurements from the Swedish and Finnish ozone measurement stations in the region.  
\end{abstract}

@@

%{\bf OBS: The following is just a copy of the 2018 dry deposition paper. It has to be substantially changed to meet the no-copycat-criteria. TODO: Rewrite in context of this article.}

%Ozone is an important trace gas for all lifeforms on Earth. Depending on the place of its occurrence it has either a positive or negative connotation. In the stratosphere, ozone absorbs most of the ultraviolet (UV)-light from the sun within the range of 100--315\,\unit{nm}, thus shielding the Earth's surface from the most harmful UV-radiation. In addition, ozone is a potent greenhouse gas in both, stratosphere and troposphere. With a radiative forcing of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, it is placed third, only surpassed by \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}.\\
%In the troposphere and in particular in ambient air, ozone is considered as a highly toxic pollutant. Since the industrial revolution, tropospheric background ozone concentrations have been increasing in the northern hemisphere \citep[Chapter 2]{IPCC2013}. In recent years, the number of episodes of peak concentrations has been, in general, decreasing in North America and Europe due to the implementation of air quality regulations \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. At the same time, fast developing countries, like e.g., China or India, saw a significant increase in ozone related air pollution. Continuously high concentrations of ambient air ozone are hazardous to the whole ecosystem. It is estimated that ozone is cause to an increase in pre-mature deaths \citep{WHO2008}, an average global loss of yield in the four major crops (wheat, rice, maize, and soybean) of about $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} as well as $7\,\unit{\%}$ loss in primary production in forestry \citep{GCB:Wittig2009,EP:Matyssek2012}. The impact of ozone damage on vegetation and agricultural plants may affect food security in the future especially in Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018} and might be an important additional feedback to climate change \citep{Nat:Sitch2007}.\\
%Elevated ozone levels at a site may originate from both, the local production of ozone from its precursors, which are transported, and from advection of ozone itself. Long-range ozone transport occurs regularly and might be most important in regions that otherwise lack precursors. Tropospheric ozone is produced in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) or volatile organic substances (VOCs -- also known as hydrocarbons) in the presents of nitrogen oxides (\chem{NO_x}). A typical reaction mechanism for \chem{CO} is sketched in the following. In a sequence of rapid reactions a peroxyl radical \chem{HO_2^\bullet} is formed through an initial reaction of \chem{CO} with a hydroxyl radical \chem{^\bullet OH}.
%Via a reaction between \chem{HO_2^\bullet} and \chem{NO}, \chem{NO_2} is formed which is then photolyzed. The resulting atomic oxygen reacts then with \chem{O_2} (and also under the presence of available co-reactants) to form an ozone molecule.

%Since ozone is highly reactive, its global mean life-time in the troposphere is roughly $22$ days but ranges between a few days in the tropical boundary layer to up to $1$ year in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}. The abundance of tropospheric ozone therefore varies, e.g., with time of the day, season, altitude, location \citep{ACP:Schnell2015}, or weather conditions in general \citep{ACP:Otero2018}. Typical concentrations of surface ozone range from $10\,\unit{ppb}$ over the tropical Pacific to $100\,\unit{ppb}$ in the downwind areas of highly emitting sources \citep[Chapter 8]{IPCC2013}. This variability poses a challenge on both, trend analysis from observations as well as validation and inter-comparison of models. From the observational side, the number of long-term observations (started before the 1950s) is limited and restricted to mainly European sites. Most of these indicate a doubling of tropospheric ozone since the 1950s \citep[Chapter 2]{IPCC2013}.

{\bf TODO:
\begin{itemize}
\item general about ozone $\rightarrow$ focus on surface ozone and in particular precursors with respect to the forest fires (\chem{CO}) and heat stress (\chem{VOCs}).
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}\\
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item[$\rightarrow$] Some info on stomatal conductance and its empirical calculation:
  \begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within the range $0-1$. They account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type $k$.
\item metrics of ozone
  \begin{itemize}
  \item[$\rightarrow$] Concentration based
    \begin{equation}
      \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone concentration $\chem{[O_3]}$ and the concentration threshold $\mathrm{x}$. A typical value for $\mathrm{x}=40\,\unit{ppb}$.
  \item[$\rightarrow$] Flux based
    \begin{equation}
      \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$.
  \end{itemize}
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak \citep{AE:Monks2000, AB:Klingberg2009, AE:Rummukainen1996} {\bf TODO: Find newer articles.}
  \end{itemize}
\end{itemize}
}

Following a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record - favorable conditions for boundary layer ozone formation.

>>>>

\begin{abstract}
  As a result of successful air quality regulations, the past two decades saw a steady reduction in episodic high summer peak concentrations of ozone over Europe and North America. At the same time, tropospheric background ozone levels kept increasing in the northern hemisphere until recently. An increased occurrence of persistent heatwaves, as one possible future scenario, generates favorable conditions for the formation of ambient air ozone. Vegetation, highly specialized to sub-arctic climate, is vulnerable to rapid environmental changes inflicted by global warming and might become more susceptible to ozone in the future. Over large parts of Europe the summer 2018 had been extraordinarily hot and dry and caused large wildfires in northern Sweden in particular. It can be regarded as a test case for future scenarios. In both 2018 and 2019, we have monitored ambient air ozone concentrations at NIBIO Environment Centre Svanhovd in Northern Norway. Ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance differed significantly, precipitation varied only to some extend. Coincidentally, we have observed ozone-induced visible injuries on clovers in the ozone garden at Svanhovd in 2018, but not in 2019.

  \emph{This may be due to differences in uptake of ozone ($\mathrm{POD_y}$) between the two years. Since DO3SE can be run on birch/beech and conifers, we would like to investigate whether the critical levels for these tree species were crossed during the two summer seasons. Ozone concentrations for the two missing weeks in July 2018 have been estimated based on the measurements from the Swedish and Finnish ozone measurement stations in the region.}  
\end{abstract}

@@

Tropospheric and in particular ambient air ozone is a highly toxic pollutant to all ecosystems. In continuously high concentration, ozone negatively affects life expectancy of humans \citep{WHO2008} and is cause to early senescence as well as programmed cell death in plants {\bf TODO citations}. On average, a global loss of yield in four major crops (wheat, rice, maize, and soybean) of $\approx 3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. Former can proof a vital threat to food security in fast developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where air pollution related to ozone has increased in the past years. At the same time, a successful implementation of air quality regulations reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}.

With a radiative forcing of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, ozone itself is a potent climate forcer, ranked third behind \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}, but in combination with its damaging effects on global vegetation it might provide an important additional feedback to climate change \citep{Nat:Sitch2007}.

Tropospheric ozone is formed in complex photochemical cycles in the presents of nitrogen oxides (\chem{NO_x}) involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as (biogenic) volatile organic substances (VOCs and BVOCs). In the absence of local pollution sources, episodes of high ozone concentrations can be attributed to long-range transport of precursors. Despite a relatively short tropospheric life-time of roughly $22\,\unit{days}$ (a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere) \citep{JGR:Stevenson2005,ACP:Young2013}, ozone itself is subject to advection over long distances. In northern Fennoscandia episodes of especially high concentrations of ambient air ozone in 2003 and 2006 have been traced back to emissions related to forest fires in southern Europe and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Boreal wildfires emit ozone precursors such as \chem{CO} ($6-13\,\unit{\%}$ relative to \chem{CO_2^{tot}}) and \chem{VOCs} ($0.5-1.5\,\unit{\%}$ relative to \chem{CO_2^{tot}}) \citep{AE:Cofer1990}.

Ozone displays an annual cycle with a distinct maximum typically in spring \citep{AE:Monks2000}. This spring peak is reached earlier in northern Fennoscandia than in more southern locations but is highly variable throughout the years \citep{AB:Klingberg2009, BER:Klingberg2019}. The cause for this might be a complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}.

\emph{Contributor: Ane}\\
As mentioned above, ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.

{\bf TODO:
\begin{itemize}
\item general about ozone $\rightarrow$ focus on surface ozone and in particular precursors with respect to the forest fires (\chem{CO}) and heat stress (\chem{VOCs}).
\item ozone induced damage in vegetation 
\item[$\rightarrow$] Some info on stomatal conductance and its empirical calculation:
  \begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within the range $0-1$. They account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type $k$.
\item metrics of ozone
  \begin{itemize}
  \item[$\rightarrow$] Concentration based
    \begin{equation}
      \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone concentration $\chem{[O_3]}$ and the concentration threshold $\mathrm{x}$. A typical value for $\mathrm{x}=40\,\unit{ppb}$.
  \item[$\rightarrow$] Flux based
    \begin{equation}
      \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$.
  \end{itemize}
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak \citep{AE:Monks2000, AB:Klingberg2009, AE:Rummukainen1996} {\bf TODO: Find newer articles.}
  \end{itemize}
\end{itemize}
}

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record - favorable conditions for boundary layer ozone formation.
